# Examining Relationships 

In Topic 1 we looked at ways to visualize data about a single variable. Now we want to learn ways to examine the relationship between *two variables*.

When we look at two variable we are interested in how one impacts the other. The **response variable** or dependent variable is the variable that gives the outcome of the study. The **explanatory variable** is the variable that claims to impact the response variable. 

::: {#exm-twovariables}
Identify what is the explanatory and response variable in the following study questions:

1. Does the amount of fertilizer affect how tall tomato plants grow?
2. Does a car's weight impact its mileage?
3. Do people who sleep more have lower stress levels?
:::

::: {.callout-important}
## Identifying Response and Explanatory
The roles aren't always clear-cut: sometimes either variable could reasonably influence the other. In these cases, the study design is key in dictating which is the explanatory and which is the response variable. 
:::

## Side-by-Side Boxplots (C -> Q)

If the explanatory variable is catagorical and the response variable is quantatative, then one way to visualize the potential differences in distribution is by using side-by-side boxplots. 

::: {#exm-systolicBP}
[Dataset.](https://www.statcrunch.com/app/?dataid=5143558){target="blank"} This dataset shows the race/ethnicity, age, and systolic BP for 360 patients. We would like to analyze if race/ethnicity impact patients BP.

*Note: this is a synthetic dataset based off US data trends. For real data, I recommend checking out [NHANES data](https://wwwn.cdc.gov/nchs/nhanes/){target="blank"} from the CDC (this data needs to be cleaned a bit though so I have not used it here).*

Below are side-by-side boxplots of systolic BP by race and ethnicity. What differences do you notice between the different racial and ethnic groups?

```{r}
#| label: fig-sbp
#| fig-cap: "Side-by-side histogram of systolic BP."
#| echo: false
#| fig-width: 6
#| fig-height: 8

# Read the data
df <- read.csv("datasets/race_systolic_bp.csv")

# Make sure race_ethnicity is treated as a factor
df$race_ethnicity <- as.factor(df$race_ethnicity)

par(mar = c(14, 4, 4, 2))  # bottom, left, top, right

boxplot(systolic_bp_mmHg ~ race_ethnicity, data = df,
        main = "Systolic Blood Pressure by Race/Ethnicity",
        ylab = "Systolic BP (mmHg)",
        col = "lightblue",
        las = 2,
        xlab = "")  

mtext("Race/Ethnicity", side = 1, line = 12)
```
:::



::: {.callout-important}
## Association does not equal causation!
There may be a **lurking variable** which affects both the explanatory and response variable. In the example above lurking variables could include: age distributions, socioeconomic status, and access to healthcare. 
:::

Another method of visualizing relationships between a catagorical explanatory variable and a quantitative response variable is by using a stacked histogram. 

```{r}
#| label: fig-stackedhist
#| fig-cap: "Stacked histogram of goals scored by different positions."
#| echo: false
#| fig-width: 6

library(ggplot2)

set.seed(42)

positions <- c("Forward", "Midfielder", "Defender")

n_forwards    <- 40
n_midfielders <- 50
n_defenders   <- 50

data <- rbind(
  data.frame(
    position = "Forward",
    goals = round(rgamma(n_forwards, shape = 3, scale = 4))
  ),
  data.frame(
    position = "Midfielder",
    goals = round(rgamma(n_midfielders, shape = 1.5, scale = 3))
  ),
  data.frame(
    position = "Defender",
    goals = round(rgamma(n_defenders, shape = 1, scale = 1.5))
  )
)

data$position <- factor(data$position, levels = positions)

ggplot(data, aes(x = goals, fill = position)) +
  geom_histogram(binwidth = 1, position = "stack", color = "white", boundary = -0.5) +
  scale_fill_manual(values = c(
    "Forward"    = "#E63946",
    "Midfielder" = "#F4A261",
    "Defender"   = "#2A9D8F"
  )) +
  labs(
    title = "Goals Scored by Players in a Season",
    subtitle = "Distribution by Position",
    x = "Goals Scored",
    y = "Number of Players",
    fill = "Position"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

Here we see that the histogram is colored by player position. From this visualization we can see that defenders typically don't score much. 

StatCrunch will let you do something similar, although it won't let you exactly stack the bars like above. In StatCrunch you can choose Graph > Histogram. Then you can select your response variable and group by your explanatory variable. Scroll all the way down to the multiple graph section and select the same X and Y axis. You can also use the rows and columns per page to change how many histograms are viewed at once. 

## Contingency Tables (C -> C)

::: {#exm-conttable}
A college polls 200 students to determine whether there is a relationship between a student's class year (freshman, sophomore, junior, or senior) and whether or not they live off campus. This data is shown in the **contingency table** below.

a) What percentage of freshmen live off campus?
b) What percentage of seniors live off campus?

| Class Year | On Campus | Off Campus | Total |
|:-----------|----------:|-----------:|------:|
| **Freshman**   | 42        | 13          | 55    |
| **Sophomore**  | 35        | 15         | 50    |
| **Junior**     | 18        | 30         | 48    |
| **Senior**     | 9         | 38         | 47    |
| **Total**  | 104  | 96    | 200 |
:::

::: {.callout-tip collapse="true"}
## Answer 

a) $(13/55)\times 100 = 23.6 \%$
b) $(38/47)\times 100 = 80.9\%$ 
:::

We can convert a contingency table to percents. To do this we convert counts to percentages when restricting to a single explanatory variable.  For example the row 

| Class Year | On Campus | Off Campus | Total |
|:------|----------:|-----------:|------:|
| **Freshman**   | 42        | 13          | 55    |

becomes

| Class Year | On Campus | Off Campus | Total |
|:------|----------:|-----------:|------:|
| **Freshman**   | $(42/55)\times 100 = \mathbf{76.4\%}$        | $(13/55)\times 100 = \mathbf{23.6\%}$          | 100%    |

In this case the rows should each add up to 100%. If our explanatory variables were in the columns, then we would make the columns sum to 100%. 

## Scatterplots (Q -> Q)

::: {#exm-study}
We want to determine if more time spent studying improves the score received on an exam. The amount of time spent studying was self reported by students. [Open the data in StatCrunch.](https://www.statcrunch.com/app/?dataid=5143819){target="blank"}

Since we want to know if study time improves exam score, the explanatory variable is study time and the reesponse variable is exam score. We can plot these using a scatter plot (Graph > Scatter Plot). We should set the study time as the x variable and the exam score as the y variable. 
:::

```{r}
#| label: fig-scatter
#| fig-cap: "Scatter plot of time spent studying and exam scores."
#| echo: false
#| fig-width: 6

df <- read.csv("datasets/student_study_scores.csv")

plot(df$study_hours,df$exam_score,
    xlab="Hours spent studying",
    ylab="Exam Score",
    pch=19)
```

We see from the scatter plot that hours spent studying is **positively correlated** with exam scores. This means that the more hours spent studying, the higher of an exam score a student tends to receive. Notice that the points appear to show a "fuzzy" line with a positive slope.

Average hours of exercise weekly is **negatively correlated** with resting heart rate. This means the more hours exercised, the lower the resting heart rate. 

```{r}
#| label: fig-scatter2
#| fig-cap: "Scatter plot of weekly exercise hours and resting heart rate."
#| echo: false
#| fig-width: 6

set.seed(42)

n <- 40

# Simulate weekly exercise hours (0 to 12 hours)
exercise_hours <- round(runif(n, 0, 12), 1)

# Resting heart rate negatively correlated with exercise, plus noise
resting_hr <- round(85 - exercise_hours * 2 + rnorm(n, 0, 6))
resting_hr <- pmin(pmax(resting_hr, 45), 100)  

# Scatter plot
plot(exercise_hours, resting_hr,
     xlab = "Average Hours Exercised per Week",
     ylab = "Resting Heart Rate (bpm)",
     pch = 19,
     col = "#2A9D8F")
```