# Examining Relationships 

In this section we want to learn ways to examine the relationship between two variables. Typically, we are asking research questions about whether one variable impacts the other. The **response variable** or dependent variable is the variable that gives the outcome of the study. The **explanatory variable** is the variable that claims to impact the response variable. 

Just like in Topic 1, these variables can be **quantitative** or **categorical**. In these notes we will spend some time looking at examples and the ways we can visualize and interpret relationships.

## Side-by-Side Boxplots (C -> Q)

If the explanatory variable is catagorical and the response variable is quantatative, then one way to visualize the potential differences in distribution is by using side-by-side boxplots. 

::: {#exm-systolicBP}
[Dataset.](https://www.statcrunch.com/app/?dataid=5143558){target="blank"} This dataset shows the race/ethnicity, age, and systolic BP for 360 patients (this data is simulated and not real). We would like to analyze if race/ethnicity impact patients BP. 

To graph side-by-side boxplots in StatCrunch first select Graph > Boxplot. Next, select the column systolic_BP_mmHg and "group by" race_ethnicity. 
```{r}
#| label: fig-sbp
#| fig-cap: "Side-by-side histogram of systolic BP."
#| echo: false
#| fig-width: 6
#| fig-height: 8

# Read the data
df <- read.csv("datasets/race_systolic_bp.csv")

# Make sure race_ethnicity is treated as a factor
df$race_ethnicity <- as.factor(df$race_ethnicity)

par(mar = c(14, 4, 4, 2))  # bottom, left, top, right

boxplot(systolic_bp_mmHg ~ race_ethnicity, data = df,
        main = "Systolic Blood Pressure by Race/Ethnicity",
        ylab = "Systolic BP (mmHg)",
        col = "lightblue",
        las = 2,
        xlab = "")  

mtext("Race/Ethnicity", side = 1, line = 12)
```
:::

Based on the graph, it looks like the race Black/African American has a higher Q1 and median than the other categories. We should be careful qualifying claims with just the above plots though! 

::: {.callout-important}
What could be other reasons the population of black patients appear to have a higher systolic BP that isn't related to their race?
:::

Another method of visualizing relationships between a catagorical explanatory variable and a quantitative response variable is by using a stacked histogram. This can be done in StatCrunch as well. 

```{r}
#| label: fig-stackedhist
#| fig-cap: "Stacked histogram of goals scored by different positions."
#| echo: false
#| fig-width: 6

library(ggplot2)

set.seed(42)

positions <- c("Forward", "Midfielder", "Defender")

n_forwards    <- 40
n_midfielders <- 50
n_defenders   <- 50

data <- rbind(
  data.frame(
    position = "Forward",
    goals = round(rgamma(n_forwards, shape = 3, scale = 4))
  ),
  data.frame(
    position = "Midfielder",
    goals = round(rgamma(n_midfielders, shape = 1.5, scale = 3))
  ),
  data.frame(
    position = "Defender",
    goals = round(rgamma(n_defenders, shape = 1, scale = 1.5))
  )
)

data$position <- factor(data$position, levels = positions)

ggplot(data, aes(x = goals, fill = position)) +
  geom_histogram(binwidth = 1, position = "stack", color = "white", boundary = -0.5) +
  scale_fill_manual(values = c(
    "Forward"    = "#E63946",
    "Midfielder" = "#F4A261",
    "Defender"   = "#2A9D8F"
  )) +
  labs(
    title = "Goals Scored by Players in a Season",
    subtitle = "Distribution by Position",
    x = "Goals Scored",
    y = "Number of Players",
    fill = "Position"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )
```

Here we see that the histogram is colored by player position. From this visualization we can see that defenders typically don't score much. 

## Contingency Tables (C -> C)

## Scatterplots (Q -> Q)

::: {#exm-study}
We want to determine if more time spent studying improves the score received on an exam. The amount of time spent studying was self reported by students. [Open the data in StatCrunch.](https://www.statcrunch.com/app/?dataid=5143819){target="blank"}

Since we want to know if study time improves exam score, the explanatory variable is study time and the reesponse variable is exam score. We can plot these using a scatter plot (Graph > Scatter Plot). We should set the study time as the x variable and the exam score as the y variable. 
:::

```{r}
#| label: fig-scatter
#| fig-cap: "Scatter plot of time spent studying and exam scores."
#| echo: false
#| fig-width: 6

df <- read.csv("datasets/student_study_scores.csv")

plot(df$study_hours,df$exam_score,
    xlab="Hours spent studying",
    ylab="Exam Score",
    pch=19)
```

We see from the scatter plot that hours spent studying is **positively correlated** with exam scores. This means that the more hours spent studying, the higher of an exam score a student tends to receive. Notice that the points appear to show a "fuzzy" line with a positive slope.

Average hours of exercise weekly is **negatively correlated** with resting heart rate. This means the more hours exercised, the lower the resting heart rate. 

```{r}
#| label: fig-scatter2
#| fig-cap: "Scatter plot of weekly exercise hours and resting heart rate."
#| echo: false
#| fig-width: 6

set.seed(42)

n <- 40

# Simulate weekly exercise hours (0 to 12 hours)
exercise_hours <- round(runif(n, 0, 12), 1)

# Resting heart rate negatively correlated with exercise, plus noise
resting_hr <- round(85 - exercise_hours * 2 + rnorm(n, 0, 6))
resting_hr <- pmin(pmax(resting_hr, 45), 100)  

# Scatter plot
plot(exercise_hours, resting_hr,
     xlab = "Average Hours Exercised per Week",
     ylab = "Resting Heart Rate (bpm)",
     pch = 19,
     col = "#2A9D8F")
```